Determinism, Complexity, and Predictability in Computer Performance

### 2014-08-04 concl.tex
rest

{\color{blue}agreed, I don't think it addds anything}\alert{I'm not sure this paragraph adds anything.  Can/should we
  remove it?  Given that information is a fundamental limitation in
  predictability, then gathering and using more information is an
  obvious next step.  But there is an equally obvious tension here
  between data length and prediction speed: a forecast that requires
  half a second to compute is not useful for the purposes of real-time
  control of a computer system with a MHz clock rate, for instance.
  Another alternative is to sample several system variables
  simultaneously and build multivariate models.  This is a particular
  challenge in nonlinear LMA-type models, since multivariate
  delay-coordinate embedding (e.g.,
  \cite{cao-multivariate-embedding,deyle-sugihara2011}) can be
  computationally prohibitive.  We are working on alternative methods
  that sidestep that complexity.}

>>>>

% {\color{blue}agreed, I don't think it addds anything}\alert{I'm not
%   sure this paragraph adds anything.  Can/should we remove it?  Given
%   that information is a fundamental limitation in predictability, then
%   gathering and using more information is an obvious next step.  But
%   there is an equally obvious tension here between data length and
%   prediction speed: a forecast that requires half a second to compute
%   is not useful for the purposes of real-time control of a computer
%   system with a MHz clock rate, for instance.  Another alternative is
%   to sample several system variables simultaneously and build
%   multivariate models.  This is a particular challenge in nonlinear
%   LMA-type models, since multivariate delay-coordinate embedding
%   (e.g., \cite{cao-multivariate-embedding,deyle-sugihara2011}) can be
%   computationally prohibitive.  We are working on alternative methods
%   that sidestep that complexity.}
%

### 2014-08-04 results.tex
rest

Figure~\ref{fig:wpe_vs_mase_best} plots the WPE values versus the
corresponding MASE values of the \emph{best} prediction for each of
the 120 time series in this study.
\begin{figure}
  \centering
  \includegraphics[width=1.1\columnwidth]{figs/new_prediction_vs_entropy}
  \caption{Weighted permutation entropy vs. mean absolute scaled error
    (MASE) of the best prediction of each time series.  The solid
    curve is a least-squares log fit of these points.
%
% of all the points except those from {\tt dgesdd$_1$}, which we have
% excluded for reasons explained in the text.
%
The dashed curves reflect the standard deviation of the model in its
parameter space.  The meaning of the shaded region is explained in the
text. \alert{[[Do you think we should explain that here, rather than
      punting to the text?  It is, after all, the primary claim of the
      paper...]]}}
  \label{fig:wpe_vs_mase_best}
\end{figure}
%
% As the plot shows, the best-case prediction error for each trace from
% these eight systems is roughly proportional to the log of the weighted
% permutation entropy, which is consistent with our first conjecture.
%
There is an obvious upward trend, which is consistent with the notion
that there is a pattern in the WPE-MASE relationship.  However, a
simple linear fit is a bad idea here.  First, any signal with zero
entropy should be perfectly predictable (i.e., MASE $\approx 0$), so
any curve fitted to these data should pass through the origin.
Moreover, WPE does not grow without bound, so one would expect the
patterns in the WPE-MASE pairs to reach some sort of asymptote.  For
these reasons, we chose to fit a function of the form $y = a \log(b x
+ 1)$ to these points\footnote{The specific values of the coefficients
  are $a=7.97 \times 10^{-2}$ and $b=1.52 \times 10^3$.}, with $y =$
WPE and $x=$ MASE.  The solid curve in the figure shows this fit; the
dashed curves show the standard deviation of this model in its
parameter space: i.e., $y = a \log(b x + 1)$ with $\pm$ one standard
deviation on each of the two parameters.  Points that fall within this
deviation volume (light grey) \alert{[[should we consider
      crosshatching or something else, given the printing
      wierdness?]]} correspond to predictions that are comparable to
the best ones found in this study; points that fall \emph{above} that
volume (dark grey) are better still.  We chose to truncate the shaded
region because of a subtle point regarding the MASE of an ideal
predictor, which should not be larger than 1 unless the training and
test signals are different.  This is discussed at more length below.

@@

\svdone---the dark blue segment of
Figure~\ref{fig:svd-ts-colored}---behaves very differently than the
other seven systems in this study.  Though its weighted permutation
entropy is very high ($0.957 \pm 0.016$), three of the four prediction
methods do quite well on this signal, yielding mean MASE scores of
0.714 (\arima), 0.827 (LMA), and 0.933 (random walk).  This pushes the
corresponding clusters of points in Figure~\ref{fig:wpe_vs_mase_all}
well above the trend followed by the other seven signals.  The reasons
for this are discussed in the following paragraph.  The MASE scores of
the predictions that were produced by the \naive ~method for this
system, however, are highly inconsistent.  The majority of the blue
diamond-shaped points on the top-right plot in
Figure~\ref{fig:wpe_vs_mase_all} are clustered near a MASE score of
0.6, which is better than the other three methods.  In five of the 15
\svdone trials, however, there were step changes in the signal.
\alert{[[Seems like same thing as we pointed out and discussed in
      \col.  We should either point out the correspondence or explain
      why it's different.]]}  The \naive ~method has a very difficult
time with signals like this, particularly if there are multiple step
changes.  This raised the MASE scores of these trials, pushing the
corresponding points to the right\footnote{This includes the cluster
  of three points near MASE $\approx 2.5$, as well as two points that
  are beyond the domain of the graph, at MASE $\approx 11.2-14.8$.},
and in turn raising both the mean and variance of this set of trials.

@@

The curves in Figures~\ref{fig:wpe_vs_mase_best}
and~\ref{fig:wpe_vs_mase_all} were determined from finite sets of
methods and data.  We put a lot of thought and effort into making
these sets representative and comprehensive.  The forecast methods
involved range from the simple to the sophisticated; the time-series
data analyzed in this section was sampled from a system whose behavior
spans the dynamical behavior space.  While we are cautiously
optimistic about the generality of our conclusions, more exploration
will be required before we can make any such claim.  Our preliminary
work along those lines shows that data from the Henon map
\cite{henon}, the Lorenz system \cite{lorenz}, the SFI A data set
\cite{weigend-book,sfi-data}, and a white-noise process all fall
within the one-$\sigma$ volume of the fit in
Figures~\ref{fig:wpe_vs_mase_best} and~\ref{fig:wpe_vs_mase_all}
region, as do various nonlinear transformations of \svdtwo, \svdfive
and \svdsix.

>>>>

Figure~\ref{fig:wpe_vs_mase_best} plots the WPE values versus the
corresponding MASE values of the \emph{best} prediction for each of
the 120 time series in this study.
\begin{figure}
  \centering
  \includegraphics[width=1.1\columnwidth]{figs/new_prediction_vs_entropy}
  \caption{Weighted permutation entropy vs. mean absolute scaled error
    (MASE) of the best prediction of each time series.  The solid
    curve is a least-squares log fit of these points.
%
% of all the points except those from {\tt dgesdd$_1$}, which we have
% excluded for reasons explained in the text.
%
The dashed curves reflect the standard deviation of the model in its
parameter space.  Points that lie below and to the right of the shaded
region indicate that the time series has more predictive structure
than the forecast strategy is able to utilize.}
  \label{fig:wpe_vs_mase_best}
\end{figure}
%
% As the plot shows, the best-case prediction error for each trace from
% these eight systems is roughly proportional to the log of the weighted
% permutation entropy, which is consistent with our first conjecture.
%
There is an obvious upward trend, which is consistent with the notion
that there is a pattern in the WPE-MASE relationship.  However, a
simple linear fit is a bad idea here.  First, any signal with zero
entropy should be perfectly predictable (i.e., MASE $\approx 0$), so
any curve fitted to these data should pass through the origin.
Moreover, WPE does not grow without bound, so one would expect the
patterns in the WPE-MASE pairs to reach some sort of asymptote.  For
these reasons, we chose to fit a function of the form $y = a \log(b x
+ 1)$ to these points\footnote{The specific values of the coefficients
  are $a=7.97 \times 10^{-2}$ and $b=1.52 \times 10^3$.}, with $y =$
WPE and $x=$ MASE.  The solid curve in the figure shows this fit; the
dashed curves show the standard deviation of this model in its
parameter space: i.e., $y = a \log(b x + 1)$ with $\pm$ one standard
deviation on each of the two parameters.  Points that fall within this
deviation volume (light grey) correspond to predictions that are
comparable to the best ones found in this study; points that fall
\emph{above} that volume (dark grey) are better still.  We chose to
truncate the shaded region because of a subtle point regarding the
MASE of an ideal predictor, which should not be larger than 1 unless
the training and test signals are different.  This is discussed at
more length below.

@@

\svdone---the dark blue segment of
Figure~\ref{fig:svd-ts-colored}---behaves very differently than the
other seven systems in this study.  Though its weighted permutation
entropy is very high ($0.957 \pm 0.016$), three of the four prediction
methods do quite well on this signal, yielding mean MASE scores of
0.714 (\arima), 0.827 (LMA), and 0.933 (random walk).  This pushes the
corresponding clusters of points in Figure~\ref{fig:wpe_vs_mase_all}
well above the trend followed by the other seven signals.  The reasons
for this are discussed in the following paragraph.  The MASE scores of
the predictions that were produced by the \naive ~method for this
system, however, are highly inconsistent.  The majority of the blue
diamond-shaped points on the top-right plot in
Figure~\ref{fig:wpe_vs_mase_all} are clustered near a MASE score of
0.6, which is better than the other three methods.  In five of the 15
\svdone trials, however, there were step changes in the signal.  This
is a different nonstationarity than in the case of \col---large jump
discontinuities rather than small shifts in the baseline---and not one
that we were able to handle by simply forcing the ARIMA model to
include a particular term.  The \naive ~method has a very difficult
time with signals like this, particularly if there are multiple step
changes.  This raised the MASE scores of these trials, pushing the
corresponding points to the right\footnote{This includes the cluster
  of three points near MASE $\approx 2.5$, as well as two points that
  are beyond the domain of the graph, at MASE $\approx 11.2-14.8$.},
and in turn raising both the mean and variance of this set of trials.

@@

The curves in Figures~\ref{fig:wpe_vs_mase_best}
and~\ref{fig:wpe_vs_mase_all} were determined from finite sets of
methods and data.  We put a lot of thought and effort into making
these sets representative and comprehensive.  The forecast methods
involved range from the simple to the sophisticated; the time-series
data analyzed in this section was sampled from a system whose behavior
spans the dynamical behavior space.  While we are cautiously
optimistic about the generality of our conclusions, more exploration
will be required before we can make any such claim.  Our preliminary
work along those lines shows that data from the H\'enon map
\cite{henon}, the Lorenz system \cite{lorenz}, the SFI A data
set\footnote{This dataset can be retrieved from
  \url{http://www-psych.stanford.edu/~andreas/Time-Series/SantaFe.html}.}
\cite{weigend-book}, and a random-walk process all fall within the
one-$\sigma$ volume of the fit in Figures~\ref{fig:wpe_vs_mase_best}
and~\ref{fig:wpe_vs_mase_all} region, as do various nonlinear
transformations of \svdtwo, \svdfive and \svdsix.

### 2014-08-04 wpe.tex
rest

Here we will be utilizing a variation of the basic permutation entropy
technique, the \emph{weighted permutation entropy} (WPE), which was
introduced in~\cite{fadlallah2013}.  \alert{We never actually say WHY
  we use it.}  The intent behind the weighting is to correct for
observational noise that is larger than the trends in the data, but
smaller than the larger-scale features.  Consider, for example, a
signal that switches between two fixed points and contains some
additive noise. The PE of such a signal will be dominated by the noise
about the fixed points, driving it to $\approx 1$---which in some
sense hides the fact that the signal is actually quite structured.  In
this situation, the terms of the weighted permutation entropy are
dominated by the switching rather than the stochastic fluctuations. To
accomplish this, the \emph{weight} of a permutation is taken into
account:
\begin{align*}
  w(x_{i+1}^\ell) = \frac{1}{\ell} \sum_{j = i + 1}^{i+\ell}
                      \left( x_j - \bar{x}_{i+1}^\ell \right)^2
\end{align*}
where $x_{i+1}^\ell$ is a sequence of values $x_{i+1}, \ldots,
x_{i+\ell}$, and $\bar{x}_{i+1}^\ell$ is the arithmetic mean of
those values.

>>>>

Here we will be utilizing a variation of the basic permutation entropy
technique, the \emph{weighted permutation entropy} (WPE), which was
introduced in~\cite{fadlallah2013}.  The intent behind the weighting
is to correct for observational noise that is larger than the trends
in the data, but smaller than the larger-scale features.  Consider,
for example, a signal that switches between two fixed points and
contains some additive noise. The PE of such a signal will be
dominated by the noise about the fixed points, driving it to $\approx
1$---which in some sense hides the fact that the signal is actually
quite structured.  In this situation, the terms of the weighted
permutation entropy are dominated by the switching rather than the
stochastic fluctuations.  The amplitude of the switches plays a key
role in predictability, so weighting the scale of ordinal changes is
important for quantifying predictive structure accurately.  To
accomplish this, WPE takes into account the \emph{weight} of a
permutation:
\begin{align*}
  w(x_{i+1}^\ell) = \frac{1}{\ell} \sum_{j = i + 1}^{i+\ell}
                      \left( x_j - \bar{x}_{i+1}^\ell \right)^2
\end{align*}
where $x_{i+1}^\ell$ is a sequence of values $x_{i+1}, \ldots,
x_{i+\ell}$, and $\bar{x}_{i+1}^\ell$ is the arithmetic mean of
those values.
